Natural selection in the Health and Retirement Study

### 2023-09-26 hrs-selection.qmd
Add discussion of true effect sizes and calculations

options(digits = 3)
set.seed(271075)
knitr::knit_hooks$set("inline", 
  \(x) {
    if (! is.na(x <- suppressWarnings(as.numeric(x)))) {
      x <- prettyNum(formatC(x, digits = 3, format = "f"), ",")
    }
    x
  }
)

library(haven)
library(forcats)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
 # masks dotchart:
suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))

nice_names <- c(
  SCZ         = "Schizophrenia",
  WC          = "Waist circ.",
  WHR         = "Waist-hip ratio",
  NEUROT      = "Neuroticism",
  WELLB       = "Well-being",
  DEPSYMP     = "Depr. symptoms",
  CD          = "Cor. art. disease",
  MI          = "Myoc. infarction",
  CORT        = "Cortisol",
  T2D         = "Type 2 diabetes",
  BIP         = "Bipolar",
  ADHD_PGC17  = "ADHD",
  XDISORDER   = "Cross-disorder",
  MENA        = "Age at menarche",
  MENO        = "Age at menopause",
  EXTRAVER    = "Extraversion",
  AUTISM      = "Autism",
  AB          = "Antisoc. behav.",
  OCD         = "OCD",
  AFBC        = "Age 1st birth combn",
  AFBF        = "Age 1st birth F",
  AFBM        = "Age 1st birth M",
  MDD2        = "Maj. depr. disorder",
  PTSDAA      = "PTSD Afr",
  PTSDEA      = "PTSD Eur",
  PTSDC       = "PTSD combined",
  HDL         = "HDL cholesterol",
  LDL         = "LDL cholesterol",
  TC          = "Total cholesterol",
  ANXFS       = "Anxiety CC", 
  ANXCC       = "Anxiety factor",
  BUN         = "Blood urea nitro.",
  BUNTE       = "Blood urea nitro. TE",
  CKD         = "Kidney disease",
  CKDTE       = "Kidney disease TE",
  DBP         = "Diastolic blood pr.",
  BMI2        = "BMI",
  HEIGHT2     = "Height",
  AI          = "Age smoking init.",
  CPD_GSCAN19 = "Cigs per day",
  DPW         = "Drinks per week",
  SC          = "Smoking cessation",
  SI          = "Smoking initiation",
  HTN         = "Hypertension",
  CANNABIS    = "Cannabis",
  GWAD2NA     = "Alzheimer's no APOE",
  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
  GWAD2WA     = "Alzheimer's",
  ALZ_01AD2WA = "Alzheimer's p 0.01",
  ALC         = "Alcohol dependence",
  PP          = "Pulse pressure",
  SBP         = "Systolic blood pr.",
  EGFR        = "Glom. filtr.",
  EGFRTE      = "Glom. filtr. TE",
  EA3         = "Educ. attainment",
  HBA1CAA     = "HbA1c Afr",
  HBA1CEA     = "HbA1c Eur",
  GCOG2       = "Gen. cognition"
)

@@

  x_short
}

@@

# rand family respondents file
# Done just once:
# rand_fam <- haven::read_dta(
#   "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
# saveRDS(rand_fam, "data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")

# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)

@@

pgs <- pgen4e |>
    rename_with(shorten) |> 
    # weird that I have to operate on whole df to use select semantics
    select(-hhid, -pn, -version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -EDU2, -EDU3, -HEIGHT, -BMI,
           -AD, -AD2, -GENCOG, -EVRSMK, -LONG, -CPD_TAG10,
           -ADHD_PGC10, -MDD
           ) |>
    names()

@@

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
         rabyear, raehsamp, raestrat, parent_weight, ragender)

@@

# clean up cache so it doesn't get too big
rm(rand_orig, trk_orig, rands, rand_fam)
```

@@

@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents (the different PGS are not a "sample" of anything).

@@

: Bootstrap estimates and 95% confidence intervals for mean absolute coefficients of PGS on RLRS in "low" and "high" groups among white respondents. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth. `r n_reps_x_es` bootstraps. {#tbl-x-bootstraps tbl-colwidths="[16,28,28,28]"}

The economic theory of fertility in @hugh2022human implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^2] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups. I tested this with bootstraps, but confidence intervals were always too wide to be informative.

[^2]: See @hugh2022human equation (6) and following.

@@

@@

>>>>

options(digits = 3)
set.seed(271075)

library(tibble)
library(haven)
library(forcats)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(carData)
library(car, warn.conflicts = FALSE) # masks some
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop
 # masks dotchart:
suppressPackageStartupMessages(library(survey, warn.conflicts = FALSE))

knitr::knit_hooks$set("inline", 
  \(x) {
    if (! is.na(x <- suppressWarnings(as.numeric(x)))) {
      x <- prettyNum(formatC(x, digits = 3, format = "f"), ",")
    }
    x
  }
)

nice_names <- c(
  SCZ         = "Schizophrenia",
  WC          = "Waist circ.",
  WHR         = "Waist-hip ratio",
  NEUROT      = "Neuroticism",
  WELLB       = "Well-being",
  DEPSYMP     = "Depr. symptoms",
  CD          = "Cor. art. disease",
  MI          = "Myoc. infarction",
  CORT        = "Cortisol",
  T2D         = "Type 2 diabetes",
  BIP         = "Bipolar",
  ADHD_PGC17  = "ADHD",
  XDISORDER   = "Cross-disorder",
  MENA        = "Age at menarche",
  MENO        = "Age at menopause",
  EXTRAVER    = "Extraversion",
  AUTISM      = "Autism",
  AB          = "Antisoc. behav.",
  OCD         = "OCD",
  AFBC        = "Age 1st birth combn",
  AFBF        = "Age 1st birth F",
  AFBM        = "Age 1st birth M",
  MDD2        = "Maj. depr. disorder",
  PTSDAA      = "PTSD Afr",
  PTSDEA      = "PTSD Eur",
  PTSDC       = "PTSD combined",
  HDL         = "HDL cholesterol",
  LDL         = "LDL cholesterol",
  TC          = "Total cholesterol",
  ANXFS       = "Anxiety factor", 
  ANXCC       = "Anxiety case-control",
  BUN         = "Blood urea nitro.",
  BUNTE       = "Blood urea nitro. TE",
  CKD         = "Kidney disease",
  CKDTE       = "Kidney disease TE",
  DBP         = "Diastolic blood pr.",
  BMI2        = "BMI",
  HEIGHT2     = "Height",
  AI          = "Age smoking init.",
  CPD_GSCAN19 = "Cigs per day",
  DPW         = "Drinks per week",
  SC          = "Smoking cessation",
  SI          = "Smoking initiation",
  HTN         = "Hypertension",
  CANNABIS    = "Cannabis",
  GWAD2NA     = "Alzheimer's no APOE",
  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
  GWAD2WA     = "Alzheimer's",
  ALZ_01AD2WA = "Alzheimer's p 0.01",
  ALC         = "Alcohol dependence",
  PP          = "Pulse pressure",
  SBP         = "Systolic blood pr.",
  EGFR        = "Glom. filtr.",
  EGFRTE      = "Glom. filtr. TE",
  EA3         = "Educ. attainment",
  HBA1CAA     = "HbA1c Afr",
  HBA1CEA     = "HbA1c Eur",
  GCOG2       = "Gen. cognition"
)

@@

  x_short <- str_replace(x_short, "^01", "ALZ_01") # two alzheimer's variants
  x_short
}

@@

# rand family respondents file
# Done just once:
# rand_fam <- haven::read_dta(
#   "data/randhrsfam1992_2018v2_STATA/randhrsfamr1992_2018v2.dta")
# saveRDS(rand_fam, "data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
rand_fam <- readRDS("data/randhrsfam1992_2018v2_STATA/rand_fam.rds")
# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h10ageokid, h10gkid, r10sbdied)

# 2010 (and others) record "age started smoking"
# rand_2010 <- haven::read_dta("data/hd10f6a_STATA/hd10f6a.dta")
# saveRDS(rand_2010, "data/hd10f6a_STATA/rand_2010.rds")
rand_2010 <- readRDS("data/hd10f6a_STATA/rand_2010.rds")
rand_2010 <- rand_2010 |> 
  select(hhid, pn, mc120, matches("mlb041")) |> 
  mutate(
    age_smoked = ifelse(mc120 %in% c(95, 98, 99), NA_real_, mc120)
  )
rand_2010$anxiety <- rand_2010 |> 
  select(matches("mlb041")) |> 
  rowMeans(na.rm = TRUE)
rand_2010$anx_na <- rand_2010 |> 
  select(matches("mlb041")) |> 
  is.na() |> 
  rowSums()
# matches the documentation at 
# https://hrs.isr.umich.edu/sites/default/files/biblio/HRS2006LBQscale.pdf :
rand_2010$anxiety[rand_2010$anx_na > 2] <- NA_real_

# 2016 survey had a module with 18 questions diagnosing adult ADHD
# rand_2016 <- haven::read_dta("data/h16f2c_STATA/h16f2c.dta")
# saveRDS(rand_2016, "data/h16f2c_STATA/rand_2016.rds")
rand_2016 <- readRDS("data/h16f2c_STATA/rand_2016.rds")
rand_2016 <- rand_2016 |> 
  select(hhid, pn, pv001:pv018) |> 
  mutate(
    across(matches("pv\\d\\d\\d"), \(x) ifelse(x %in% 8:9, NA, x)),
    across(matches("pv\\d\\d\\d"), \(x) x == 1) # 1 is "yes', 5 is "no"
  )
rand_2016$adhd_score <- rand_2016 |> 
  select(pv001:pv018) |> 
  rowSums(na.rm = TRUE)
rand_2016$adhd_nas <- rand_2016 |> 
  select(pv001:pv018) |> 
  is.na() |> 
  rowSums(na.rm = TRUE)
rand_2016$adhd_score <- rand_2016$adhd_score/(18 - rand_2016$adhd_nas)

@@

old_pgs <- c("EDU2", "EDU3", "HEIGHT", "BMI",
           "AD", "AD2", "GENCOG", "EVRSMK", "LONG", "CPD_TAG10",
           "ADHD_PGC10", "MDD")
pgs <- pgen4e |>
    rename_with(shorten) |> 
    # weird that I have to operate on whole df to use select semantics
    select(-hhid, -pn, -version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -any_of(old_pgs)
           ) |>
    names()

@@

rand <- rand |> left_join(rand_2016, 
                          by = join_by(hhid, pn),
                          unmatched = "drop",
                          relationship = "one-to-one")

rand <- rand |> left_join(rand_2010, 
                          by = join_by(hhid, pn),
                          unmatched = "drop",
                          relationship = "one-to-one")

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

rand$neurot <- rand |> 
  select(matches("r\\d+lbneur")) |> 
  rowMeans(na.rm = TRUE)

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(old_pgs), 
         all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
         rabyear, raehsamp, raestrat, parent_weight, ragender,
         adhd_score, neurot, r10diabe, r10hearte, r10cesd, age_smoked,
         anxiety, r10cogtot)

@@

# clean up cache so it doesn't get too big
rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
```

@@

@fig-rlrs-edu plots each PGS's partial correlation with RLRS against its partial correlation with educational attainment. The relationship is negative among whites (correlation `r cors_fert_edu$white[1]`, bootstrap 95% C.I. `r ci_boot_fert$white[1]` to `r ci_boot_fert$white[2]`) and negative but insignificant among blacks (correlation `r cors_fert_edu$black[1]`, bootstrap 95% C.I. `r ci_boot_fert$black[1]` to `r ci_boot_fert$black[2]`). Survey bootstraps [@canty1999resampling] are used so as to make inferences from the sample of respondents.

@@

: Bootstrap estimates and 95% confidence intervals for mean absolute coefficients of PGS on RLRS in "low" and "high" groups among white respondents. Groups are: 0-12 years education vs. 13-17 years; below vs. above median income; all others vs. married; below vs. above median age at first birth. `r n_reps_x_es` bootstraps. {#tbl-x-bootstraps tbl-colwidths="\[16,28,28,28\]"}

The economic theory of fertility also implies that correlations between education and RLRS should be more negative for lower-income/education people, single parents and people who have children earlier.[^1] This in turn implies that correlations between PGS coefficients on education and on RLRS should be more negative among these groups. I tested this with bootstraps, but confidence intervals were always too wide to be informative.

[^1]: See @hugh2022human equation (6) and following.

@@

Lastly, we would like to know natural selection's effect sizes. The bivariate correlations of PGS with RLRS gives the change in one generation in the PGS due to natural selection, measured in standard deviations. Polygenic scores contain error, so estimated correlations are biased towards zero compared to the correlation of the true PGS. They can be scaled up by

$$
\hat{\beta}_{TRUE} = \hat{\beta}_{PGS}\sqrt{\frac{h^2}{R^2_{PGS}}
}$$

where $h^2$ is the heritability of the PGS target phenotype and $R^2_{PGS}$ is the $R^2$ of the measured PGS on the target phenotype [@becker2021resource]. I make this calculation using heritabilities from twin studies, and SNP- or chip-heritabilities, reported in @abdellaoui2021dissecting (chip-heritability is the total variance explained by common SNPs recorded by array chips).

@fig-true-effects reports estimates for the true PGS as well as the "true chip-PGS" - the true best predictor of the phenotype using only variants recorded on the array chip, for phenotypes where selection effects were nominally significant and a measure of the phenotype was available in-sample and was significantly predicted by the PGS. It's important to understand that estimates for the true PGS are more uncertain than for the true chip-PGS. Ideally, the true chip-PGS differs only from the recorded PGS because of sampling error in calculating the underlying summary statistics.[^2] The true PGS would use many more genetic variants, perhaps including e.g. *de novo* mutations, and the relationship between these variants' effect on fertility and on the target phenotype may not be the same as for genetic variants captured on array chips.

[^2]: But even the chip-PGS might contain non-random noise from correlated environments.

The HRS contains multiple PGS for several phenotypes, typically with the $R^2$ increasing over time. Appendix @fig-check-correction runs a sanity check in which I predict the coefficient of later scores on fertility, using earlier scores and the above formula. Mostly, earlier PGS do a good job at predicting the effect of later PGS on fertility, with the later PGS's point estimate within the earlier PGS's 95% confidence interval. One exception is that the point estimate of the earlier PGS for general cognition, GENCOG, on the effect of the later PGS GCOG2 has the wrong sign.

As a rough rule of thumb, a 0.1 standard deviation change in a polygenic score over a generation might count as "serious": about 54% of the new generation will be below the parents' mean. Many upper confidence bounds meet that threshold, but lower bounds are often small or include zero. Given the uncertainties about the exact nature of "noise" in existing PGS, the estimates -- even for chip heritability -- mostly show the limits of our knowledge, and should be treated as best guesses only.

```{r}
#| label: stat-boot-true-effects

#' Calculate corrected PGS
#'
#' @param w Weights
#' @param data Data frame
#' @param pgs Name(s) of polygenic score(s) columns in  `data`.
#' @param phenotype Name of target phenotype in `data`
#' @param chip_h2 Chip-heritability
#' @param twin_h2 Twin-study heritability
#'
#' @return A vector of corrected estimates of effects on fertility, 
#'   named like `source.to.target` 
#'   where `source` is one of `pgs` and target is either a `pgs` or the
#'   chip-PGS or true PGS. If `pgs` has more than one element then
#'   the function will correct pgs using each other as a sanity check. If
#'   `source == target` then that is the original uncorrected score.
#' @noRd
#'
#' @examples
boot_correct <- function (w, data, pgs, phenotype, chip_h2, twin_h2) {

  fs_fert <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
  fs_phen <- map(pgs, \(x) reformulate(c(x, pcs), response = phenotype))

  ms_fert <- map(fs_fert, \(x) lm(x, data = data, weights = w))
  ms_phen <- map(fs_phen, \(x) lm(x, data = data, weights = w))
  bhat_fert <- map2_dbl(ms_fert, pgs, \(x, y) coef(x)[[y]])
  bhat_phen <- map2_dbl(ms_phen, pgs, \(x, y) coef(x)[[y]])

  r2_phen <- map_dbl(ms_phen, \(x) {
    an <- car::Anova(x)
    an$`Sum Sq`[1]/sum(an$`Sum Sq`)  # incremental R2 of first term
  })

  r2_phen <- c(r2_phen, chip_h2, twin_h2)
  # the correction factor
  # lower triangular part of this shows how much you multiply to get from
  # COLNAME to ROWNAME
  rhos <- outer(r2_phen, r2_phen, \(x, y) sqrt(x/y)) 
  rownames(rhos) <- colnames(rhos) <- c(pgs, "Chip", "Twin")

  # multiply first column by bhat_edu[1] to give predictions from the
  # first PGS to the next ones; etc.
  # The diagonals of b_edu are exactly equal to bhat_edu
  # arow,column gives the predicted value of b by ROWNAME for COLNAME
  b_phen <- apply(rhos, 1, \(x) x * c(bhat_phen, NA, NA))
  b_fert <- apply(rhos, 1, \(x) x * c(bhat_fert, NA, NA))

  # don't include the last two rows of NA predictions "from chip/twin h2"
  b_fert <- b_fert[seq_along(pgs), , drop = FALSE]

  res <- c(b_fert)
  # TODO: are the "backward estimates" wrong?
  # TODO: check against the python tool from the PGI repository paper
  res_names <- outer(rownames(b_fert), colnames(b_fert),
                      \(x, y) paste0(x, ".to.", y))
  names(res) <- c(res_names)

  res
}

n_reps_true <- 99
rand_boot <- rand |> as.svrepdesign(type = "bootstrap", 
                                    replicates = n_reps_true)

correct_scores <- tibble::tribble(
  ~pgs                         , ~phenotype  , ~chip_h2, ~twin_h2,
  c("ADHD_PGC10", "ADHD_PGC17"), "adhd_score", 0.21    , 0.75    ,
  c("EDU2", "EDU3", "EA3")     , "raedyrs"   , 0.10    , 0.43    ,  
  "AFBC"                       , "agefbn"    , 0.05    , 0.26    ,
  "NEUROT"                     , "neurot"    , 0.10    , 0.37    ,
  "CD"                         , "r10hearte" , 0.22    , 0.4     ,
  "T2D"                        , "r10diabe"  , 0.18    , 0.22    , 
  c("MDD", "MDD2")             , "r10cesd"   , 0.08    , 0.37    ,
#  "ANXCC"                      , "anxiety"   , 0.31    , 0.45    ,
  "AI"                         , "age_smoked", 0.05    , 0.50    ,
  c("GENCOG", "GCOG2")         , "r10cogtot" , 0.19    , 0.86
)

correct_scores <- drop_na(correct_scores)
# Sources for chip/twin h2; Abdel.
# general cognition used h2 for "adult IQ"
# Except:
# heart disease 0.4 to 0.6 from https://www.nejm.org/doi/full/10.1056/nejm199404143301503
# via https://www.jacc.org/doi/full/10.1016/j.jacc.2014.12.033
# T2D  0.22 to 0.73...
# https://www.cell.com/ajhg/pdf/S0002-9297(12)00031-6.pdf
# https://www.cambridge.org/core/services/aop-cambridge-core/content/view/94E0005D626FB0399AA8FDDAC82F970E/S1832427415000833a.pdf/div-class-title-the-concordance-and-heritability-of-type-2-diabetes-in-34-166-twin-pairs-from-international-twin-registers-the-discordant-twin-discotwin-consortium-div.pdf
# anxiety (nb not social anxiety!)
# https://ajp.psychiatryonline.org/doi/full/10.1176/ajp.156.2.246
# generalized anxiet disorder 42%:
# https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0134865
# age of smoking initiation, including chip heritability
# http://www.matthewckeller.com/publications/Evans_Addiction_2021.pdf
# CAD chip heritability
# https://academic.oup.com/cardiovascres/article/113/8/973/2965332?login=false
# type ii diabetes chip heritability
# https://www.nature.com/articles/s41588-018-0241-6
# 'current anxiety symptoms' chip heritability
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7237282/

results_df <- list()
for (i in seq(nrow(correct_scores))) {
  source_pgs <- correct_scores$pgs[[i]]
  phenotype <- correct_scores$phenotype[[i]]
  chip_h2 <- correct_scores$chip_h2[[i]]
  twin_h2 <- correct_scores$twin_h2[[i]]

  results <- withReplicates(rand_boot, theta = boot_correct, 
                            pgs = source_pgs, phenotype = phenotype, 
                            chip_h2 = chip_h2, twin_h2 = twin_h2)
  results_df[[phenotype]] <- results |>
    as.data.frame() |>
    tibble::rownames_to_column(var = "rn") |>
    cbind(confint(results)) |>
    tibble::remove_rownames() |>
    setNames(c("rn", "mean", "SE", "conf.low", "conf.high")) |> 
    mutate(
      phenotype = phenotype,
      source.pgs = str_remove(rn, ".to.*"),
      target.pgs = str_remove(rn, ".*.to."),
      rn = NULL,
      .before = 1
    )
}

results_df <- list_rbind(results_df)
```

```{r}
#| label: fig-true-effects
#| fig-cap: |
#|   Estimated effects on RLRS of true polygenic scores calculated
#|   using estimates of chip- and twin-heritability and the most recent
#|   polygenic score.

# TODO: maybe just use the polygenic score name not the phenotype?
# TODO: why are some result bounds negative? e.g. ADHD? Presumably
#. because can have negative effect on RLRS?
# TODO: check the original phenotypes from which h2 ests were derived...
#. - is CAD the same as coronary heart disease (r10hearte)?
# TODO: why is ANXCC not predicting anxiety?
results_df |> 
  mutate(
    Phenotype = nice_names[source.pgs],
    Heritability = target.pgs
  ) |> 
  filter(Heritability %in% c("Chip", "Twin"), ! source.pgs %in% old_pgs) |> 
  ggplot(aes(y = Phenotype, x = mean, xmin = conf.low, xmax = conf.high,
             color = Heritability)) +
    geom_vline(xintercept = 0, color = "grey60") +
    geom_pointrange(aes(linetype = Heritability == "Twin"),
                    position = position_dodge(0.2)) +
    scale_linetype(guide = "none") +
    coord_cartesian(xlim = c(- 0.3, 0.3)) +
    labs(x = "Corrected effect on RLRS", y = "") +
    theme_minimal()

@@

```{r}
#| label: fig-check-correction
#| fig-cap: |
#|   Estimated effects of series of PGS using errors-in-variables 
#|   correction. "Own" is the uncorrected estimate using the PGS
#|   itself. Other estimates are from earlier or later PGS targeting
#|   the same phenotype and corrected via the method in @becker2021resource.
#| fig-height: 6

results_df |> 
  filter(length(unique(source.pgs)) > 1, .by = phenotype) |> 
  filter(! target.pgs %in% c("Chip", "Twin")) |> 
  mutate(
    Phenotype = case_match(phenotype,
                           "adhd_score" ~ "ADHD",
                           "r10cesd" ~ "Depression",
                           "r10cogtot" ~ "Cognition",
                           "raedyrs" ~ "Educ. attainment"
                           ),
    Target = fct_relevel(target.pgs,
                         "ADHD_PGC17", "ADHD_PGC10", "GCOG2", "GENCOG",
                         "MDD2", "MDD", "EA3", "EDU3", "EDU2"),
    Source = ifelse(source.pgs == target.pgs, "Own", source.pgs),
  ) |> 
  ggplot(aes(y = Target, x = mean, xmin = conf.low, xmax = conf.high,
             color = Source)) +
    geom_vline(xintercept = 0, color = "grey60") +
    geom_pointrange(position = position_dodge(0.7)) +
    geom_text(aes(x = mean, label = Source), hjust = "center", 
              vjust = 1.5,
              position = position_dodge(0.7), size = 3) + 
    scale_color_manual(values = c("Own" = "red"), guide = "none") +
    facet_grid(rows = vars(Phenotype), scales = "free", space = "free") +
    labs(x = "Estimated effect on RLRS with correction", y = "Target PGS") +
    coord_cartesian(xlim = c(-0.05, 0.05)) + 
    theme_minimal()
